---
title: 'powerLATE: Generalized Power Analysis for LATE'
output:
  html_document: default
  pdf_document: default
---

***

This page demonstrates the usage of the **powerLATE** package. 

***

**Author:** [Kirk Bansak](https://www.kirkbansak.com/) (UCSD); [Eddie Yang](https://www.eddieyang.net/) (UCSD); 

**Date:** August 25, 2020

**Version:** 0.1.0

**Reference:** [Bansak, K. (2020). A Generalized Approach to Power Analysis for Local Average Treatment Effects](https://projecteuclid.org/download/pdfview_1/euclid.ss/1591171230)

R code used in this demonstration can be downloaded from [here](https://github.com/EddieYang211/powerLATE_aux/blob/master/code.R).

***

# Contents

1. Examples using `powerLATE()`

2. Examples using `powerLATE.cov()`

***

# 1. Examples using `powerLATE()`
### 1.1 Recovering power

Suppose we are running an experiment in which we expect some noncompliance with the treatment assignment and hence want to estimate the local average treatment effect (LATE). We are interested in knowing what power we will have to reject the null hypothesis that the LATE equals zero, so we load the powerLATE package to perform power calculations.

```{r}
library(powerLATE)
```

Now suppose in our experiment that exactly half of the experimental subjects will be assigned to treatment and the other half to control. Further suppose the sample size is 3000, the compliance rate is 0.5, and the LATE effect size is 0.25. Without making the ordered-means assumption, we can compute our power by the following:

```{r}
results <- powerLATE(pi = 0.5, N = 3000, kappa = 0.25)
```

***

In addition to the output that is printed in the console, `powerLATE()` returns both the list of input parameters and output parameters as invisible objects. We can get these values by typing:

```{r}
results$input.parameter
results$output.parameter
```

***

We can also supply a list of compliance rates to get their corresponding power. Additionally, the treatment-assignment probability, `pZ` can be unequal (!= 0.5). Here we set `pZ = 0.25`
```{r}
powerLATE(pZ = 0.25, pi = c(0.3,0.4,0.5,0.6,0.7), N = 3000, kappa = 0.25)
```

***

We can also make the ordered-means assumption by setting `assume.ord.means = TRUE`, this will result in a higher conservative (lower) bound for power. (Note: please check that your data indeed comforms to the ordered-means assumption. For details on the assumption, see [Bansak (2020)](https://projecteuclid.org/download/pdfview_1/euclid.ss/1591171230))

```{r}
powerLATE(pi = c(0.3,0.4,0.5,0.6,0.7), N = 3000, kappa = 0.25, assume.ord.means = TRUE)
```

***

If the absolute effect is preferred to the effect size, we can recover power by supplying `tau` and `omega`, and setting `effect.size = FALSE`.
```{r}
powerLATE(pi = 0.5, N = 3000, effect.size = FALSE, tau = 300, omega = 1500)
```

***

### 1.2 Recovering required sample size

In addition to making the ordered-means assumption and specifying absolute effect, if we fix `power` to be 0.8 but instead want to know the required sample sizes for various compliance rates, we can do so by
```{r}
powerLATE(pi = c(0.5,0.6,0.7,0.8), power = 0.8, assume.ord.means = TRUE,
          effect.size = FALSE, tau = 25, omega = 125)
```

***

### 1.3 Recovering minimum detectable effect size

To recover minimum detectable effect size, simply supply `power` and `N` (sample size). Here we supply a list of sample sizes and setting the compliance rate `pi = 0.6` and `power = 0.8`
```{r}
powerLATE(pi = 0.6, N = c(1000,1500,2000,2500,3000), power = 0.8)
```

***

### 1.4 Recovering minimum detectable effect (absolute)

Instead of minimum detectable effect size, we can instead get the minimum detectable absolute effect by setting `effect.size = FALSE` and supplying `omega`.
```{r}
powerLATE(pZ = 0.4, pi = 0.6, N = c(1000,1500,2000,2500,3000),
          power = 0.8, assume.ord.means = TRUE, 
          effect.size = FALSE, omega = 50)
```

***

# 2. Examples using `powerLATE.cov()`
The syntax and arguments of `powerLATE.cov()` are identical to `powerLATE()`, except that `powerLATE.cov()` requires two additional arguments:

* `r2dw`: proportion of variation in D left unexplained by Z that is explained by covariates. 
* `r2yw`: proportion of variation in Y left unexplained by Z that is explained by covariates.

Using the scenario in 1.4, we can calculate the minimum detectable absolute effect with `powerLATE.cov()` by
```{r}
powerLATE.cov(pZ = 0.4, pi = 0.6, N = c(1000,1500,2000,2500,3000),
              power = 0.8, assume.ord.means = TRUE, 
              effect.size = FALSE, omega = 50, r2dw = 0.15, r2yw = 0.10)
```

Notice the conservative (upper) bounds for `tau` are slightly lower than those in 1.4, as part of the variation in treatment assignment and outcome is soaked up by covariate adjustment, resulting in a more powerful estimator. 

***

Please report bugs and let us know if you have any suggestions! -> z5yang [at] ucsd.edu



